EGGs are adhesive!

### 2025-03-14 calco/main.tex
update conclusioni

Note also that for their nature, all the rules adopted for EGGs have the identity as the left-hand side. 
Potentially, they thus belong to any choice of $\mathcal{M}$, even if the drawback is the 
possibility to keep on repeating the same rewriting step.
However, this is easily forbidden adopting rules \emph{with negative application conditions}, which are given by the usual span 
$L \leftarrow K \rightarrow R$ plus an additional arrow $n: L\rightarrow N$, such that a match $m: L \to G$ is admissible if it cannot
be factorised through $n$. For EGGs and rules $L \leftarrow L \rightarrow R$, $n$ is simply the right-hand side itself.
%
The theory of $\mathcal{M}$-adhesivity carries on with such rules, see~\cite{ehrig2012,ehrig2014adhesive}.

\section{Conclusions and further and related works}
\label{conclusioni}
The aim of our paper was to extend the theory of $\mathcal{M}$-adhesive categories in order to include EGGS, 
a formalism for program optimisation 
%via a compact representation and 
%efficient implementation of equality saturation.
%
To do so, we revisited the notion of hyper-graph and introduced  term graph with equivalence, 
proving that they are $\mathcal{M}$-adhesive categories, and we extended this results in order to capture
also prove the same property for EGGs as term graphs with equivalence satisfying a suitable closure constraint.
Summing up, we proved that EGGs are objects of an $\mathcal{M}$-adhesive category $\eg$ 
and that optimisation steps are obtained via DPO rules (possibly with negative application conditions) whose 
left-hand side is in $\mathcal{M}$, and in fact an identity. 
%and via a match that is in $\mathcal{N}$. 
This fact allows for exploiting the properties 
of the $\mathcal{M}$-adhesive framework when discussing about the optimisation process.

@@

\emph{Related works.}
%\todo{espandere un poco}
Despite the interest they have ben raising as an efficient data structure, we are not aware of any attempt to provide an 
algebraic characterisation of EGGs, the only exception being~\cite{ghica}. We leave for future work an in-depth comparison 
among the two proposals, which appear to be related yet quite different.
%
In both cases, the key intuition is  to use string diagrams to represents (acyclic) term graphs. We adopted a
more down-to-Earth approach by equipping
concretely the nodes of a term graph with an equivalence relation, thus directly equivalent to the original presentation~\cite{DetlefsNS05},
at the same time extending and generalising~\cite{concur2006} in the contemporary jargon of adhesive categories.
The route chosen in~\cite{ghica} is to consider term graphs as arrows of a symmetric monoidal category,
and equipping them with an enrichment over semi-lattices, the resulting formalism being reminiscent of hierarchical hypergraphs.
Thus, the solution in~\cite{ghica}  is  more general than ours since it can be lifted to term graphs defined over categories other than
$\Set$.
%
At the same time, we both consider the DPO approach for rewriting, even if only our solution guarantees that the resulting category
is actually $\mathcal{M}$-adhesive, thus allowing to exploit all the features of the framework
as they hold for standard graph transformation.

>>>>

Note also that for their nature, all the rules adopted for EGGs have the identity as the left-hand side. 
Potentially, they thus belong to any choice of $\mathcal{M}$, even if the drawback is the 
possibility to keep on repeating the same rewriting step.
However, this is easily forbidden adopting rules \emph{with negative application conditions}, given as the usual span 
$L \leftarrow K \rightarrow R$ plus an additional arrow $n: L\rightarrow N$, such that a match $m: L \to G$ is admissible if it cannot
be factorised through $n$~\cite{HabelHT96}. For EGGs and rules $L \leftarrow L \rightarrow R$, $n$ is simply the right-hand side itself.
%
The theory of $\mathcal{M}$-adhesivity carries on with such rules, see~\cite{ehrig2012,ehrig2014adhesive}.

\section{Conclusions and further and related works}
\label{conclusioni}
The aim of our paper was to extend the theory of $\mathcal{M}$-adhesive categories in order to include EGGS, 
a formalism for program optimisation 
%via a compact representation and 
%efficient implementation of equality saturation.
%
To do so, we revisited the notions of hyper-graphs and term graphs with equivalence, 
proving that they are $\mathcal{M}$-adhesive categories, and we extended these results in order to
prove the same property for EGGs as term graphs with equivalence satisfying a suitable closure constraint.
Summing up, we proved that EGGs are objects of an $\mathcal{M}$-adhesive category $\eg$ 
and that optimisation steps are obtained via DPO rules (possibly with negative application conditions) whose 
left-hand side is in $\mathcal{M}$,
%, and in fact an identity. 
%and via a match that is in $\mathcal{N}$. 
and fact allows for exploiting the properties 
of the $\mathcal{M}$-adhesive framework.
% when discussing about the optimisation process.

@@

\emph{Related works.}
%\todo{espandere un poco}
Despite the interest they have ben raising as an efficient data structure, we are not aware of any attempt to provide an 
algebraic characterisation of EGGs, the only exception being~\cite{ghica}. We leave for future work an in-depth comparison 
among the two proposals, which appear to be related yet quite different.
%
In both cases, the key intuition is  to use string diagrams to represents term graphs. We adopted a
more down-to-Earth approach by equipping
concretely the nodes of a term graph with an equivalence relation, thus directly corresponding to the original presentation~\cite{DetlefsNS05},
at the same time extending and generalising it~\cite{concur2006} in the contemporary jargon of adhesive categories.
The route chosen in~\cite{ghica} is to consider term graphs as arrows of a symmetric monoidal category,
and equipping them with an enrichment over semi-lattices, the resulting formalism being reminiscent of hierarchical hypergraphs.
Thus, the solution in~\cite{ghica}  is  more general than ours since it can be lifted to term graphs defined over categories other than
$\Set$.
%
At the same time, we both consider the DPO approach for rewriting, even if only our solution guarantees that the resulting category
is actually $\mathcal{M}$-adhesive, thus allowing to exploit all the features of the framework
as they hold for standard graph transformation.
